Learning how to act: making good decisions with machine learning

### 2017-09-20 thesis/thesis_map.tex
addresses Tiberio's comments

\section{Motivation}
Many of the most important questions in science, commerce and our personal lives are about the outcomes of doing something. Will asking people to pay upfront at the doctors reduce long term health expenditure? If we developed a drug to suppress particular genes, could we cure multiple sclerosis and would delaying teen-aged pregnancies improve the outcome for their children.

These are hard questions because they require more than identifying a pattern in data. Correlation is not causation. Causal inference has proven so difficult that there is barely any consensus on even enduring questions like the returns to education or the long-term consequences of early life events – like teenage pregnancy, despite the fact that the variables involved are susceptible to human intuition and understanding.

@@

The notion of causality has been widely debated in science and philosophy \citep{hume1741treatise,mill1893system,pearson1911grammar,Pearl2000,Lewis2000,woodward2005making,heckman20051,cartwright2007hunting} but is still widely viewed as poorly defined. This has led to a reluctance among applied researchers in many fields to make concrete claims about causality in their work, leading them instead to report that variables are \textit{related}, \textit{correlated} or \textit{associated}. However, the magnitude, direction and even existence of an association depends which variables are controlled for (or included in a regression). Avoiding formalising causation, which is the real question of interest, requires the reader to determine via \quotes{common sense} the implications of the reported associations.

@@

Causality is often liked to explanation; understanding how and why things happen. I view explanation in terms of compression and generalisabilty: the amount of information about the world a model can capture. This creates a hierarchy in the degree to which models are explanatory, rather than a simple binary distinction. A standard predictive model encodes all the information needed to predict some output given inputs provided the system generating the data does not change. A high-level causal model might be able to predict the outcome of a specific intervention holding all else fixed. More detailed causal models could predict the outcome for a wide range of combinations of interventions conditional on a range of contexts. By considering conditional interventions within our definition of causal questions we also capture mediation: the study of pathways through which one variable causes another \cite{vanderweele2015explanation}. Finally, a model that can distil how elements interact into mathematical equations like Newton's laws can be used to predict what will happen in an astounding range of settings, including many never previously observed.\footnote{Although Newton's laws are not fully general, as they break down for extremely massive objects}

@@

There has traditionally been a large gap between researchers in machine learning who focus on prediction, using largely non-interpretable models and researchers in statistics, social science and economics who (at least implicitly) aim to answer causal questions and tend to use highly theory-driven models. However, there is relatively little awareness, particularly within the machine learning and data science communities, of what constitutes a causal problem and the implications of this for the training and evaluation models. In the next section we emphasise the subtly that can exist in determining if a problem is causal by examining some typical examples.

@@

Machine learning is in the midst of a boom, driven by the availability of large data sets and the computation resources to process them. Machine learning techniques are being applied a huge range of problems, in both industry and academia. The following examples are intended to capture the breadth of problems that machine learning algorithms are actively being applied to. Which, if any, of these problems require causal inference?

@@

What about predicting the risk of death in patients with pneumonia? Suppose the goal is to build a model to decide who should be treated in hospital and who can be sent home with antibiotics. If we assume that in hospital treatment is more effective for serious cases, this appears to be straightforward prediction. It is not. Depending on how the decision to admit was previously made and what features are included (or omitted) in the model, the relationship between those features and the outcome may change if the model is used to make admission decisions. \citet{Caruana2015} found exactly this effect in a real data set. The model learned that people suffering asthma were \emph{less} likely to die from pneumonia. This was because doctors treated such patients very aggressively, thus actually lowering their risk. The issue is not with model; it performed very well at the task for which it was trained, which is to predict who would be likely to die under the original admission and treatment protocols. However, using it to decide how to \emph{change} these protocols could kill. The actual question of interest in this case is what happens to patients with characteristics $X$ they are assigned treatment according to decision rule (or policy) $\pi(X)$.

@@

Predicting which customers will leave or who will re-offend if granted parole also fit within the category of problems where you wish to identify a group for which a problem will occur and target some treatment to them (hospitalisation, loyalty reward, more monitoring \& support on parole, etc). Predictive models can be applied to such problems provided it is known which treatment option will be most effective for the target group and deciding who to treat on the based of the model predictions won't change the relationship between features and outcome.

@@

The problem of playing go is causal from the interventionalist perspective. The agent wishes to learn the probability of a win given an action they take. However, there are some special characteristics to the go problem that make it amenable to a primarily supervised learning approach. The actions the agent has to explore are the sames ones as human players explored to generate the training data, and both have the same objective - to win the game. In addition, the state of the board encapsulates all the information relevant to selecting a move. These factors make it reasonable to conclude that selecting moves with an algorithm will not change the value of a board state or the probability of given move by the opponent by a sufficiently large margin to invalidate the training data.

@@

As we have shown, problems involving causality are ubiquitous in many fields. As a result, techniques for addressing them have developed in parallel within many disciplines, including statistics, economics, social science, epidemiology and machine learning. Although the focus and terminology can differ substantially between fields these techniques, all address the underlying goal of estimating the effect of, or optimally selecting, interventions in some system. Methods for learning about interventions can be usefully categorised into two broad approaches, reinforcement learning and observational causal inference.

In reinforcement learning, under which we include traditional randomised experiments, we learn the outcome of actions by taking them. We take the role of an agent, capable of intervening in the system, and aim to develop algorithms that allow the agent to select actions optimally with respect to some goal. A particular strand of research within reinforcement learning are multi-armed bandit problems. They describe settings in which there are a set of available actions, the agent repeatedly decides which to select and then observes the outcome of the chosen action. They capture problems such as a doctor deciding which treatment to prescribe to a patient or a search engine selecting which advertisement to display to a user, where the agent faces the same set of choices repeatedly and is able to assess the value of outcome of the actions.

The approach of learning the outcome of an action by taking it plays a key role in advancing our knowledge of the world. However, we frequently have access to large bodies of data that have been collected from a system in which we did not have any control over what actions were taken, or perfect knowledge of the basis on which those actions were chosen. Estimating the effect of an action from such observational data sets is the problem addressed by observational causal inference. Observational causal inference can be viewed as a form of transfer learning. The goal is to leverage data obtained from one system, the system we have observed, to estimate key characteristics of another, the system after we select an action via some policy that may differ from the process driving which actions occur in the original system. This is impossible without some assumptions about how the the original system and the system after intervention are related to one-another. The key to observational inference is to model how actions change the state of the world in such a way that we can map information collected in one setting to another.

Both multi-armed bandits and observational causal inference can be seen as extensions to the concept of randomised controlled trials. Bandit algorithms deal with the sequential nature of the decision making process, and causal inference with problems where randomisation is not feasible, affordable or ethical. The similarities between the problems addressed by these techniques raises the question of whether there are problems best addressed by a combination of these approaches, and if so, how they can be combined.

@@

\paragraph{Contributions} The goal of this thesis is to connect and unify the key approaches to solving causal problems from both the observational and interventional viewpoints. My major contribution is a framework that unifies the causal graphical model approach for inference in observational settings with the sequential experimental approach encapsulated by multi-armed bandits. This framework allows us to represent knowledge of how variables are related to one-another in a very natural way and induces an interesting and novel form of structure between the different actions modelled in the bandit problem. I develop a new algorithm that can exploit this structure as a first step towards a unified approach to decision making under uncertainty. 
I make a number of additional connective contributions that are not encompassed by my work on causal bandit problems. I demonstrate the role of a formal causal framework within Bayesian approaches to inference and show how assigning a prior based on human causal intuition without considering the causal structure of the problem can introduce bias. I highlight the connections between approaches to off-policy evaluation in bandit problems, causal effect estimation from observational data, and covariate shift. I clarify the implicit causal structure underlying various bandit settings and the counterfactual nature of \emph{regret} - the measure by which bandit algorithms are assessed.

\paragraph{Thesis overview} This thesis is divided into three key chapters, learning from observational data, learning from interventional data and unifying the approaches. Chapter \ref{chap:causality} covers learning to act from observational data. In \S\ref{sec:causal_models}, I describe the key existing frameworks for causal inference from observational data, discuss how they relate to one-another and introduce the notation required to describe causal problems. Section \ref{sec:causal effect estimation} describes the key tools these frameworks provide that enable us to answer causal questions, in particular, the do-calculus (\S\ref{sec:do_calculus}) and, in sections \ref{sec:defining-causal-effects} and \ref{sec:causal_effect_under_ignorability}, discusses how we define causal effects, the traditional approaches to estimation and how they relate to covariate shift and off-policy evaluation. Section \ref{sec:bayesian_causality} highlights the role graphical causal models can play in Bayesian inference.

Chapter \ref{chap:interventionalist_view} deals with interventionalist viewpoint, including traditional randomised experiments (\S\ref{sec:randomized_experiment}) and multi-armed bandit problems. In \S\ref{sec:multi-armed-bandits}, I describe the key problems and results within the bandit literature, including stochastic bandits (\S\ref{sec:stochastic-bandit}), pure exploration problems (\S\ref{sec:pure-exploitation-problems}), adversarial bandits (\S\ref{sec:adversarial-bandits}) and contextual bandits (\S\ref{sec:contextual-bandits}). I clarify the causal structure of (stochastic) contextual bandit problems in \S\ref{sec:causal-structure-contextual-bandit}. In \S\ref{sec:off-policy-evaluation}, I review the literature on off-policy evaluation for bandit problems and show how it is a somewhat special case of causal effect estimation from observational data. Finally, in section \S\ref{sec:counterfactual-nature-regret} I discuss the counterfactual nature bandit regret.

In chapter \ref{chap:causal-bandit}, I introduce causal bandit problems: a framework that unifies causal graphical models with multi-armed bandit problems. Bandit arms are related to interventions in a causal graphical model in a very natural way: each multi-armed bandit arm (or action) corresponds to a particular assignment of values to variables within the causal graphical model. I show how the framework can be used to describe a number of existing problems that lie in the intersection between the observational and interventional approaches to causality and demonstrate how problems reduce to different flavours of bandit problem depending what information is observable and when.

In \S\ref{sec:causal-bandit-post-action-feedback}, I focus on causal bandit problems for which the values of variables in the causal graph are observed after an action is selected. I demonstrate that this leads to a novel form of structure between the bandit arms that cannot be exploited by any existing bandit algorithms. In \S\ref{sec:simple-regret}, I describe and develop an algorithm for a special case of causal bandit problem I refer to as the \emph{parallel bandit problem}. I demonstrate via upper and lower bounds on the regret that the algorithm is close to optimal for these problems and that the introduction of the causal structure leads to substantially faster learning. In \S\ref{sec:simple-regret-general}, I develop and prove regret bounds for an algorithm that can be applied to arbitrary causal graphs, albeit with stronger assumptions on what must be known apriori, and introduce a measure that captures the underlying difficulty of causal bandit problems, which depends on the causal graph and can be viewed as an \quotes{effective number of arms}. Section \ref{sec:causal-bandit-experiments} provides experiments demonstrating the performance of these algorithms on causal bandit problems. Finally in \S\ref{sec:causal_bandit_discussion}, I discuss extensions and potential future work.

@@

The goal of causal inference is to learn the effect of taking an action. We can do this directly via experimental approaches, however any given agent only has a limited capacity to manipulate the world. We are generating and storing data on almost every aspect of our lives at an unprecedented rate. As we incorporate sensors and robotics into our cities, homes, cars, everyday products and even our bodies, the breadth and scale of this data will only increase. However, only a tiny fraction of this data will be generated in a controlled way with the specific goal of answering a single question. An agent that can only learn from data when it had explicit control (or perfect knowledge of) the process by which that data was generated will be severely limited. This makes it critical that we develop effective methods that enable us to predict the outcome of an intervention in some system from observing it. This is the problem of observational causal inference.

@@

Observational causal inference aims to infer the outcome of an intervention in some system from data obtained by observing (but not intervening in) it. As previously mentioned, this is a form of transfer learning; we need to infer properties of the system post-intervention from observations of the system pre-intervention. Mapping properties from one system to another requires some assumptions about how these two systems are related, or other words, a way of describing actions and how we anticipate a system will respond to them. Three key approaches have emerged: counterfactuals, structural equation models and causal Bayesian networks.

Counterfactuals \citep{Rubin1974} were developed from the starting point of generalising from randomised trials to less controlled settings. They describe causal effects in terms of differences between counterfactual variables, what would happen if we took one action verses what would happen if we took another. Counterfactual assertions can be expressed very naturally in human languages and are  prevalent in  everyday conversations; \quotes{if I had worked harder I would have got better grades} and \quotes{she would have been much sicker if she hadn't taken antibiotics}. Structural equation models have been developed and applied primarily within economics and related disciplines. They can be seen as an attempt to capture key aspects of the people's behaviour with mathematics. Questions around designing policies or interventions play a central role in economics. Thus they have transformed simultaneous equations into a powerful (though somewhat poorly formalised \citep{Pearl2000}) framework and associated set of methods for estimating causal effects. Causal Bayesian networks \citep{Pearl2000} are a more recent development and arise from the addition of a fundamental assumption about the meaning of a link to Bayesian networks. They inherit and leverage the way Bayesian networks encode conditional independencies between variables to localise the impact of an intervention in a system in a way that allows formalisation of the conditions under which causal effects can be inferred from observational data.

An understanding of causal Bayesian networks and their properties (in particular the do calculus, see section \ref{sec:do_calculus}) is sufficient to appreciate my main technical contributions in chapter \ref{chap:causal-bandit}, as well as the importance of formal causal reasoning in Bayesian inference that I highlight in section  \ref{sec:bayesian_causality}. However, the literature on causal inference techniques remains spit between the different frameworks. Much of the recent work on estimating causal effects within machine learning, as well as widely used methodologies such as propensity scoring, are described using the counterfactual framework. Methods developed within economics, in particular instrumental variable based approaches, or those requiring parametric or functional assumptions, are often based around structural equation models. This makes it worthwhile for researchers interested in causality to develop an understanding off all these viewpoints.

In the next sections, we describe causal Bayesian networks, counterfactuals and structural equation models: the problems they allow us to solve, the assumptions they rely on and how they differ. By describing all three frameworks, how they relate to one-another, and when they can be viewed as equivalent, we will make it easier for researchers familiar with one framework to understand the others and to transfer ideas and techniques between them. However, sections \ref{sec:SEM} (Structural Equation models), and \ref{sec:unifying_causal_models} (Unifying the models) are not crucial to understanding my technical contributions and may be safely skipped by uninterested readers. In order to demonstrate the notation and formalisms each framework provides, we will use them to describe the following simple examples.

@@

Causal Bayesian networks are Bayesian networks, so results that apply to Bayesian networks carry directly across: the local Markov property states that variables are independent of their non-effects given their direct causes. The global Markov property and d-separation also hold in causal networks. D-separation, which characterises which conditional independencies must hold in any distribution that can represented by a given Bayesian network $G$, is key to many important results and algorithms for causal inference. We include a brief review of D-separation in section \ref{sec:dsep}.

@@

The Neyman-Rubin model \citep{Rubin1974,Rubin1978,Rosenbaum1983, Rubin2005,Rubin2008} defines causality in terms of potential outcomes, or counterfactuals. Counterfactuals are statements about imagined or alternate realities, are prevalent in everyday language and may play a role in the development of causal reasoning in humans \citep{Weisberg2013}. Causal effects are differences in counterfactual variables: what is the difference between what would have happened if we did one thing versus what would have happened if we did something else.

@@

\begin{equation}
\begin{aligned}
&Y^{0} \equiv P(Y|do(X=0)) \\
&Y^{1} \equiv P(Y|do(X=1))
\end{aligned}
\end{equation}

@@

\subsection{Careful with that prior}

@@

\vspace{0.5cm}
\begin{theorem}
\label{theorem:cb-intervene-all-parents}
Let $\mathcal{A'}$ be the set of all possible assignments of values to the parents of $Y$. If $\mathcal{A'} \subseteq \mathcal{A}$ and $C(a') \leq C(a) \; \forall (a' \in \mathcal{A'}, a \in \mathcal{A}/\mathcal{A'})$ then the optimal action $a^* \in \mathcal{A'}$ and the problem reduces to a standard multi-armed bandit (over actions in in $\mathcal{A'}$).
\begin{proof}
for any action $a \in \mathcal{A}$, 
\eq{
 \E{Y|\vec{X}^c_t,a} &= \Esub{\parents{Y}\sim \P{\parents{Y}|\vec{X}^c_t,a}}{\E{Y|\vec{X}^c_t,a,\parents{Y}}} \\
 &= \Esub{\parents{Y}\sim \P{\parents{Y}|\vec{X}^c_t,a}}{\E{Y|\parents{Y}}}\\
 &= \Esub{\parents{Y}\sim \P{\parents{Y}|\vec{X}^c_t,a}}{\E{Y|do(\parents{Y})}}\\
 &\leq \max_{\parents{Y}}\E{Y|do(\parents{Y})} = \E{Y|a'} \text{ for some $a' \in \mathcal{A'}$}
}
This proves the optimal action $a^* \in \mathcal{A'}$

@@

A better understanding of causality is also relevant for the discussions around transparency and ethics in machine learning, particularly with respect to the European Union's new General Data Protection Regulation, which requires that automated decision systems that significantly affect individuals provide provide \textit{"meaningful information about the logic involved."} \cite{Goodman2016}. The recognition that there is a fundamental trade-off between accuracy and transparency, unless we can build perfect causal models, when the interests of individuals and society diverge has implications for they way we design and regulate systems that have the potential to have major impacts on people's lives. We must develop approaches to ensuring machine learning decisions are reasonable and ethical that allow effected individuals recourse to dispute or improve outcomes but do not undermine the ability of the system to function.

The observational and interventional viewpoints on learning to act contribute complementary components to a general approach. Observational causal inference provides, through the do-calculus, a formal means to map information from observational to interventional settings, as well as between different interventions. Bandit algorithms, capture the sequential nature of decision making processes and provide techniques to carefully balance exploration and exploitation.

I have developed a framework that formally connects causal graphical models with bandit problems in a very natural way and demonstrated that this framework encode some key existing problems in the literature. I showed that knowledge of the causal structure (but not the functional relationships) between variables can induce a novel form of structure between alternate actions and that an algorithm that leverages this structure obtains better performance that one that does not.

>>>>

\section{Motivation}
Many of the most important questions in science, commerce and our personal lives are about the outcomes of doing something. Will asking people to pay upfront at the doctors reduce long term health expenditure? If we developed a drug to suppress particular genes, could we cure multiple sclerosis? Would delaying teen-aged pregnancies improve the outcome for their children.

These are hard questions because they require more than identifying a pattern in data. Correlation is not causation. Causal inference has proven so difficult that there is barely any consensus on even enduring questions like the returns to education or the long-term consequences of early life events – like teenage pregnancy - despite the fact that the variables involved are susceptible to human intuition and understanding.

@@

The notion of causality has been widely debated in science and philosophy \citep{hume1741treatise,mill1893system,pearson1911grammar,Pearl2000,Lewis2000,woodward2005making,heckman20051,cartwright2007hunting} but is still widely viewed as poorly defined. This has led to a reluctance among applied researchers in many fields to make concrete claims about causality in their work, leading them instead to report that variables are \textit{related}, \textit{correlated} or \textit{associated}. However, the magnitude, direction and even existence of an association depends on which variables are controlled for (or included in a regression). Avoiding formalising causation, which is the real question of interest, requires the reader to determine via \quotes{common sense} the implications of the reported associations.

@@

Causality is often linked to explanation; understanding how and why things happen. I view explanation in terms of compression and generalisation: the amount of information about the world a model can capture. This creates a hierarchy in the degree to which models are explanatory, rather than a simple binary distinction. A standard predictive model encodes all the information needed to predict some output given inputs provided the system generating the data does not change. A high-level causal model might be able to predict the outcome of a specific intervention holding all else fixed. More detailed causal models could predict the outcome for a wide range of combinations of interventions conditional on a range of contexts. By considering conditional interventions within our definition of causal questions we also capture mediation: the study of pathways through which one variable causes another \cite{vanderweele2015explanation}. Finally, a model that can distil how elements interact into mathematical equations like Newton's laws can be used to predict what will happen in an astounding range of settings, including many never previously observed.\footnote{Although Newton's laws are not fully general.}

@@

There has traditionally been a large gap between researchers in machine learning who focus on prediction, using largely non-interpretable models and researchers in statistics, social science and economics who (at least implicitly) aim to answer causal questions and tend to use highly theory-driven models. However, there is relatively little awareness, particularly within the machine learning and data science communities, of what constitutes a causal problem and the implications of this for the training and evaluation of models. In the next section we emphasise the subtlety that can exist in determining if a problem is causal by examining some typical examples.

@@

Machine learning is in the midst of a boom, driven by the availability of large data sets and the computation resources to process them. Machine learning techniques are being applied to a huge range of problems, in both industry and academia. The following examples are intended to capture the breadth of problems that machine learning algorithms are actively being applied to. Which, if any, of these problems require causal inference?

@@

What about predicting the risk of death in patients with pneumonia? Suppose the goal is to build a model to decide who should be treated in hospital and who can be sent home with antibiotics. If we assume that in hospital treatment is more effective for serious cases, this appears to be straightforward prediction. It is not. Depending on how the decision to admit was previously made and what features are included (or omitted) in the model, the relationship between those features and the outcome may change if the model is used to make admission decisions. \citet{Caruana2015} found exactly this effect in a real data set. The model learned that people suffering asthma were \emph{less} likely to die from pneumonia. This was because doctors treated such patients very aggressively, thus actually lowering their risk. The issue is not with the model; it performed very well at the task for which it was trained, which is to predict who would be likely to die under the original admission and treatment protocols. However, using it to decide how to \emph{change} these protocols could kill. The actual question of interest in this case is what happens to patients with characteristics $X$ if they are assigned treatment according to decision rule (or policy) $\pi(X)$.

@@

@@

The problem of playing go is causal from the interventionalist perspective. The agent wishes to learn the probability of a win given an action they take. However, there are some special characteristics of the go problem that make it amenable to a primarily supervised learning approach. The actions the agent has to explore are the same ones as human players explored to generate the training data, and both have the same objective - to win the game. In addition, the state of the board encapsulates all the information relevant to selecting a move. These factors make it reasonable to conclude that selecting moves with an algorithm will not change the value of a board state or the probability of given move by the opponent by a sufficiently large margin to invalidate the training data.

@@

As we have shown, problems involving causality are ubiquitous in many fields. As a result, techniques for addressing them have developed in parallel within many disciplines, including statistics, economics, social science, epidemiology and machine learning. Although the focus and terminology can differ substantially between fields, these techniques all address the underlying goal of estimating the effect of, or optimally selecting, interventions in some system. Methods for learning about interventions can be usefully categorised into two broad approaches, reinforcement learning and observational causal inference.

In reinforcement learning, under which we include traditional randomised experiments, we learn the outcome of actions by taking them. We take the role of an agent, capable of intervening in the system, and aim to develop algorithms that allow the agent to select actions optimally with respect to some goal. A particular strand of research within reinforcement learning are multi-armed bandit problems. They describe settings in which there is a set of available actions, the agent repeatedly decides which to select and then observes the outcome of the chosen action. They capture problems, such as a doctor deciding which treatment to prescribe to a patient or a search engine selecting which advertisement to display to a user, where the agent faces the same set of choices repeatedly and is able to assess the value of the outcome they observe.

The approach of learning the outcome of an action by taking it plays a key role in advancing our knowledge of the world. However, we frequently have access to large bodies of data that have been collected from a system in which we did not have any control over what actions were taken, or perfect knowledge of the basis on which those actions were chosen. Estimating the effect of an action from such observational data sets is the problem addressed by observational causal inference. Observational causal inference can be viewed as a form of transfer learning. The goal is to leverage data obtained from one system, the system we have observed, to estimate key characteristics of another, the system after we select an action via some policy that may differ from the process driving which actions occur in the original system. This is impossible without some assumptions about how the original system and the system after intervention are related to one-another. The key to observational inference is to model how actions change the state of the world in such a way that we can map information collected in one setting to another.

Both multi-armed bandits and observational causal inference can be seen as extensions to the concept of randomised controlled trials. Bandit algorithms deal with the sequential nature of the decision making process, and causal inference with problems where randomisation is not feasible, affordable or ethical. The similarities between the problems addressed by these techniques raise the question of whether there are problems best addressed by a combination of these approaches, and if so, how they can be combined.

@@

\paragraph{Contributions} The goal of this thesis is to connect and unify the key approaches to solving causal problems from both the observational and interventional viewpoints. My major contribution is a framework that unifies the causal graphical model approach for inference in observational settings with the sequential experimental approach encapsulated by multi-armed bandits. This framework allows us to represent knowledge of how variables are related to one-another in a very natural way and induces an interesting and novel form of structure between the different actions modelled in the bandit problem. I develop a new algorithm that can exploit this structure as a first step towards a unified approach to decision making under uncertainty.

I also make a number of additional connective contributions that are not encompassed by my work on causal bandit problems. I demonstrate the role of a formal causal framework within Bayesian approaches to inference and show how assigning a prior based on human causal intuition without considering the causal structure of the problem can introduce bias. I highlight the connections between approaches to off-policy evaluation in bandit problems, causal effect estimation from observational data, and covariate shift. Finally, I clarify the implicit causal structure underlying various bandit settings and the counterfactual nature of \emph{regret} - the measure by which bandit algorithms are assessed.

\paragraph{Thesis overview} This thesis is divided into three key chapters: learning from observational data, learning from interventional data and unifying the approaches. Chapter \ref{chap:causality} covers learning to act from observational data. In \S\ref{sec:causal_models}, I describe the key existing frameworks for causal inference from observational data, discuss how they relate to one-another and introduce the notation required to describe causal problems. Section \S\ref{sec:causal effect estimation} describes the key tools these frameworks provide that enable us to answer causal questions, in particular, the do-calculus (\S\ref{sec:do_calculus}) and, in sections \S\ref{sec:defining-causal-effects} and \S\ref{sec:causal_effect_under_ignorability}, discusses how we define causal effects, the traditional approaches to estimation and how they relate to covariate shift and off-policy evaluation. Section \S\ref{sec:bayesian_causality} highlights the role graphical causal models can play in Bayesian inference.

Chapter \ref{chap:interventionalist_view} deals with the interventionalist viewpoint, including traditional randomised experiments (\S\ref{sec:randomized_experiment}) and multi-armed bandit problems. In \S\ref{sec:multi-armed-bandits}, I describe the key problems and results within the bandit literature, including stochastic bandits (\S\ref{sec:stochastic-bandit}), pure exploration problems (\S\ref{sec:pure-exploitation-problems}), adversarial bandits (\S\ref{sec:adversarial-bandits}) and contextual bandits (\S\ref{sec:contextual-bandits}). I clarify the causal structure of (stochastic) contextual bandit problems in \S\ref{sec:causal-structure-contextual-bandit}. In \S\ref{sec:off-policy-evaluation}, I review the literature on off-policy evaluation for bandit problems and show how it is a somewhat special case of causal effect estimation from observational data. Finally, in \S\ref{sec:counterfactual-nature-regret} I discuss the counterfactual nature of bandit regret.

In chapter \ref{chap:causal-bandit}, I introduce causal bandit problems: a framework that unifies causal graphical models with multi-armed bandit problems. Bandit arms are related to interventions in a causal graphical model in a very natural way: each multi-armed bandit arm (or action) corresponds to a particular assignment of values to variables within the causal graphical model. I show how the framework can be used to describe a number of existing problems that lie in the intersection between the observational and interventional approaches to causality and demonstrate how problems reduce to different flavours of the bandit problem depending what information is observable and when.

In \S\ref{sec:causal-bandit-post-action-feedback}, I focus on causal bandit problems for which the values of variables in the causal graph are observed after an action is selected. I demonstrate that this leads to a novel form of structure between the bandit arms that cannot be exploited by any existing bandit algorithms. In \S\ref{sec:simple-regret}, I describe and develop an algorithm for a special case of the causal bandit problem, which I refer to as the \emph{parallel bandit problem}. I demonstrate via upper and lower bounds on the regret that the algorithm is close to optimal for these problems and that the introduction of the causal structure leads to substantially faster learning. In \S\ref{sec:simple-regret-general}, I develop and prove regret bounds for an algorithm that can be applied to arbitrary causal graphs, albeit with stronger assumptions on what must be known apriori, and introduce a measure that captures the underlying difficulty of causal bandit problems, which depends on the causal graph and can be viewed as an \quotes{effective number of arms}. Section \ref{sec:causal-bandit-experiments} provides experiments demonstrating the performance of these algorithms on causal bandit problems. Finally in \S\ref{sec:causal_bandit_discussion}, I discuss extensions and potential future work.

@@

The goal of causal inference is to learn the effect of taking an action. We can do this directly via experimental approaches, however any given agent only has a limited capacity to manipulate the world. We are generating and storing data on almost every aspect of our lives at an unprecedented rate. As we incorporate sensors and robotics into our cities, homes, cars, everyday products and even our bodies, the breadth and scale of this data will only increase. However, only a tiny fraction of this data will be generated in a controlled way with the specific goal of answering a single question. An agent that can only learn from data when it had explicit control (or perfect knowledge of) the process by which that data was generated will be severely limited. This makes it critical that we develop effective methods that enable us to predict the outcome of an intervention in some system by observing, rather than acting on it. This is the problem of observational causal inference.

@@

Observational causal inference aims to infer the outcome of an intervention in some system from data obtained by observing (but not intervening on) it. As previously mentioned, this is a form of transfer learning; we need to infer properties of the system post-intervention from observations of the system pre-intervention. Mapping properties from one system to another requires some assumptions about how these two systems are related, or in other words, a way of describing actions and how we anticipate a system will respond to them. Three key approaches have emerged: counterfactuals, structural equation models and causal Bayesian networks.

Counterfactuals \citep{Rubin1974} were developed from the starting point of generalising from randomised trials to less controlled settings. They describe causal effects in terms of differences between counterfactual variables, what would happen if we took one action versus what would happen if we took another. Counterfactual assertions can be expressed very naturally in human languages and are  prevalent in  everyday conversations; \quotes{if I had worked harder I would have got better grades} and \quotes{she would have been much sicker if she hadn't taken antibiotics}. Structural equation models have been developed and applied primarily within economics and related disciplines. They can be seen as an attempt to capture key aspects of the people's behaviour with mathematics. Questions around designing policies or interventions play a central role in economics. Thus they have transformed simultaneous equations into a powerful (though somewhat poorly formalised \citep{Pearl2000}) framework and associated set of methods for estimating causal effects. Causal Bayesian networks \citep{Pearl2000} are a more recent development and arise from the addition of a fundamental assumption about the meaning of a link to Bayesian networks. They inherit and leverage the way Bayesian networks encode conditional independencies between variables to localise the impact of an intervention in a system in a way that allows formalisation of the conditions under which causal effects can be inferred from observational data.

An understanding of causal Bayesian networks and their properties (in particular the do calculus, see section \ref{sec:do_calculus}) is sufficient to appreciate my main technical contributions in chapter \ref{chap:causal-bandit}, as well as the importance of formal causal reasoning in Bayesian inference that I highlight in section  \ref{sec:bayesian_causality}. However, the literature on causal inference techniques remains split between the different frameworks. Much of the recent work on estimating causal effects within machine learning, as well as widely used methodologies such as propensity scoring, are described using the counterfactual framework. Methods developed within economics, in particular instrumental variable based approaches, or those requiring parametric or functional assumptions, are often based around structural equation models. This makes it worthwhile for researchers interested in causality to develop an understanding off all these viewpoints.

In the next sections, we describe causal Bayesian networks, counterfactuals and structural equation models: the problems they allow us to solve, the assumptions they rely on and how they differ. By describing all three frameworks, how they relate to one-another, and when they can be viewed as equivalent, we will make it easier for researchers familiar with one framework to understand the others and to transfer ideas and techniques between them. However, sections \ref{sec:SEM} (Structural Equation models), and \ref{sec:unifying_causal_models} (Unifying the models) are not crucial to understanding my technical contributions and may be safely skipped. In order to demonstrate the notation and formalisms each framework provides, we will use them to describe the following simple examples.

@@

Causal Bayesian networks are Bayesian networks, so results that apply to Bayesian networks carry directly across: the local Markov property states that variables are independent of their non-effects given their direct causes. The global Markov property and d-separation also hold in causal networks. D-separation, which characterises which conditional independencies must hold in any distribution that can be represented by a given Bayesian network $G$, is key to many important results and algorithms for causal inference. We include a brief review of D-separation in section \ref{sec:dsep}.

@@

The Neyman-Rubin model \citep{Rubin1974,Rubin1978,Rosenbaum1983, Rubin2005,Rubin2008} defines causality in terms of potential outcomes, or counterfactuals. Counterfactuals are statements about imagined or alternate realities, are prevalent in everyday language and may play a role in the development of causal reasoning in humans \citep{Weisberg2013}. Causal effects are differences in counterfactual variables: what the difference is between what would have happened if we did one thing versus what would have happened if we did something else.

@@

\begin{equation}
\begin{aligned}
&\P{Y^{0}} \equiv P(Y|do(X=0)) \\
&\P{Y^{1}} \equiv P(Y|do(X=1))
\end{aligned}
\end{equation}

@@

\subsection{Be careful with that prior}

@@

\vspace{0.5cm}
\begin{theorem}
\label{theorem:cb-intervene-all-parents}
Let $\mathcal{A'}$ be the set of all possible assignments of values to the parents of $Y$. If $\mathcal{A'} \subseteq \mathcal{A}$ and $C(a') \leq C(a) \; \forall (a' \in \mathcal{A'}, a \in \mathcal{A}/\mathcal{A'})$ then the optimal action $a^* \in \mathcal{A'}$ and Algorithm \ref{alg:general} cannot do better than uniform exploration over $\mathcal{A'}$.
\begin{proof}
for any action $a \in \mathcal{A}$, 
\eq{
 \E{Y|\vec{X}^c_t,a} &= \Esub{\parents{Y}\sim \P{\parents{Y}|\vec{X}^c_t,a}}{\E{Y|\vec{X}^c_t,a,\parents{Y}}} \\
 &= \Esub{\parents{Y}\sim \P{\parents{Y}|\vec{X}^c_t,a}}{\E{Y|\parents{Y}}}\\
 &= \Esub{\parents{Y}\sim \P{\parents{Y}|\vec{X}^c_t,a}}{\E{Y|do(\parents{Y})}}\\
 &\leq \max_{\parents{Y}}\E{Y|do(\parents{Y})} = \E{Y|a'} \text{ for some $a' \in \mathcal{A'}$}
}
This proves the optimal action $a^* \in \mathcal{A'}$

@@

A better understanding of causality is also relevant for the discussions around transparency and ethics in machine learning, particularly with respect to the European Union's new General Data Protection Regulation, which requires that automated decision systems that significantly affect individuals provide \textit{"meaningful information about the logic involved."} \cite{Goodman2016}. The recognition that there is a fundamental trade-off between accuracy and transparency, unless we can build perfect causal models, when the interests of individuals and society diverge has implications for the way we design and regulate systems that have the potential to have major impacts on people's lives. We must develop approaches to ensuring machine learning decisions are reasonable and ethical that allow effected individuals recourse to dispute or improve outcomes but do not undermine the ability of the system to function.

The observational and interventional viewpoints on learning to act contribute complementary components to a general approach. Observational causal inference provides, through the do-calculus, a formal means to map information from observational to interventional settings, as well as between different interventions. Bandit algorithms capture the sequential nature of decision making processes and provide techniques to carefully balance exploration and exploitation.

I have developed a framework that formally connects causal graphical models with bandit problems in a very natural way and demonstrated that this framework encodes some key existing problems in the literature. I showed that knowledge of the causal structure (but not the functional relationships) between variables can induce a novel form of structure between alternate actions and that an algorithm that leverages this structure obtains better performance that one that does not.

This work represents an important first step towards a unified approach to causal inference and optimal decision making. There is much exciting work remaining to be done. Although the causal bandit framework can capture contextual information as well as post-action feedback, I have formally analysed and developed algorithms only for the latter. Additionally, to make the problem tractable, I made the (major) assumption that the interventional distribution over the parents of the outcome was known. This can be relaxed to assuming the interventional distribution over some Markov blanket with respect to the outcome is known. Information can then be shared between actions outside the blanket in the same way as in algorithm \ref{alg:general}, whilst actions inside could be learned explicitly. Relaxing this assumption entirely is a much more challenging problem. However, as is demonstrated by the specific example of the parallel causal bandit problem, it is possible to develop algorithms that require only the causal structure of the graph and yield substantially lower regret.
